---
title: "Gene Set Enrichment Analysis (GSEA): Whole Body Hyperthermia (WBH) Treatment"
author: "Victoria Chen, Anisha Kolambe, Ethan Aidam"
date: "Presented Monday, May 11"
output:
  pdf_document:
    toc: true
    highlight: tango
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo        = TRUE,
  message     = FALSE,
  warning     = FALSE,
  fig.align   = "center",
  fig.width   = 8,
  fig.height  = 5,
  out.width   = "85%",
  dpi         = 150
)
```

## Overview

This analysis starts from the STAR-derived raw gene count matrix (`wbh_counts_matrix.csv`) and walks through quality filtering, differential expression with DESeq2, and a custom implementation of Gene Set Enrichment Analysis (GSEA). We test four biologically motivated gene sets, heat shock/proteostasis, interferon signaling, serotonin, and dopamine, for coordinated expression changes between Whole Body Hyperthermia (WBH)-treated and sham-treated samples.

## Load libraries

```{r load-libraries}
library(AnnotationDbi)
library(org.Hs.eg.db)
library(GO.db)
library(edgeR)
library(DESeq2)
library(tidyverse)
library(gt)
```

## Load raw STAR count matrix

Column names are cleaned of BAM and tab-file suffixes added by STAR, and Ensembl gene IDs are mapped to HGNC symbols using `org.Hs.eg.db`. Genes with no matching symbol retain their Ensembl ID, and all row names are made unique before downstream use.

```{r load-counts}
cnt <- read.csv(
  "rnaseq/count/wbh_counts_matrix.csv",
  row.names = 1,
  check.names = FALSE
)
cnt <- as.matrix(cnt)

# Clean sample names if STAR/BAM suffixes are present.
colnames(cnt) <- basename(colnames(cnt))
colnames(cnt) <- sub("\\.Aligned\\.sortedByCoord\\.out\\.bam$", "", colnames(cnt))
colnames(cnt) <- sub("\\.ReadsPerGene\\.out\\.tab$", "", colnames(cnt))

# Remove version numbers if present (e.g., ENSG... .1)
ensembl_ids <- sub("\\..*$", "", rownames(cnt))

gene_annot <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = ensembl_ids,
  columns = c("SYMBOL"),
  keytype = "ENSEMBL"
)

# Match annotation back to count matrix
gene_symbols <- gene_annot$SYMBOL[match(ensembl_ids, gene_annot$ENSEMBL)]

# Replace missing symbols with Ensembl IDs
gene_symbols[is.na(gene_symbols)] <- ensembl_ids[is.na(gene_symbols)]

# Ensure uniqueness (required for DESeq2)
gene_symbols <- make.unique(gene_symbols)
rownames(cnt) <- gene_symbols

# Sanity checks
if (any(is.na(cnt))) {
  stop("NA values detected in count matrix.")
}
if (any(cnt < 0)) {
  stop("Negative values detected. DESeq2 requires non-negative raw counts.")
}

# Inspect dimensions and first columns
dim(cnt)
head(cnt[, 1:min(4, ncol(cnt))])
```

## Define Sham versus WBH groups

Each of the 19 samples are assigned to either the WBH group (10 samples) or the Sham group (9 samples) using a manually curated sample table keyed by SRR accession. `Sham` is set as the reference level throughout. A design matrix with an intercept and a `groupWBH` indicator is created for modeling.

```{r define-groups}
sample_info <- tibble(
  sample = c(
    "SRR34413163", "SRR34413164", "SRR34413165", "SRR34413166", "SRR34413167",
    "SRR34413168", "SRR34413169", "SRR34413170", "SRR34413171", "SRR34413172",
    "SRR34413174", "SRR34413175", "SRR34413176", "SRR34413177",
    "SRR34413178", "SRR34413179", "SRR34413180", "SRR34413181", "SRR34413182"
  ),
  group = c(
    rep("WBH", 10),
    rep("Sham", 9)
  )
) %>%
  mutate(group = factor(group, levels = c("Sham", "WBH"))) %>%
  as.data.frame()

rownames(sample_info) <- sample_info$sample

# Match count columns to sample metadata.
missing_samples <- setdiff(sample_info$sample, colnames(cnt))
extra_samples   <- setdiff(colnames(cnt), sample_info$sample)

if (length(missing_samples) > 0) {
  stop(paste("Samples in sample_info but missing from cnt:",
             paste(missing_samples, collapse = ", ")))
}
if (length(extra_samples) > 0) {
  message("Samples in cnt but not sample_info will be dropped: ",
          paste(extra_samples, collapse = ", "))
}

cnt <- cnt[, sample_info$sample]

# create a group level vector
group  <- sample_info$group

# create design matrix
design <- model.matrix(~ group)

group
design
```

## Filter genes by CPM

Lowly expressed genes are removed by retaining only those with a count-per-million (CPM) greater than 1 in at least 2 samples. This step eliminates genes that are essentially absent from the data  A `DGEList` object is then constructed and TMM normalization factors are computed.

```{r filter-genes}
cpm_vals     <- cpm(cnt)
keep         <- rowSums(cpm_vals > 1) >= 2
cnt_filtered <- cnt[keep, ]

filter_summary <- tibble(
  genes_before_filtering = nrow(cnt),
  genes_after_filtering  = nrow(cnt_filtered),
  genes_removed          = nrow(cnt) - nrow(cnt_filtered)
)
filter_summary

# Store counts and phenotype labels together
dge <- DGEList(
  counts  = cnt_filtered,
  group   = sample_info$group,
  samples = sample_info
)

# Add library sizes and normalization factors
dge <- calcNormFactors(dge)

# Confirm that phenotype labels are stored
dge$samples
```

## Run DESeq2

DESeq2 is used to fit a negative-binomial model for each gene with `group` as the sole covariate, contrasting WBH against Sham.

```{r run-deseq2}
dds <- DESeqDataSetFromMatrix(
  countData = cnt_filtered,
  colData   = sample_info,
  design    = ~ group
)

# Ensure Sham is the reference level.
dds$group <- relevel(dds$group, ref = "Sham")
dds       <- DESeq(dds)

res <- results(
  dds,
  contrast = c("group", "WBH", "Sham")
)

summary(res)

res_df <- as.data.frame(res) %>%
  rownames_to_column("gene_id") %>%
  arrange(padj)

head(res_df)
```

## Create ranked gene list for GSEA by log2 fold change

GSEA requires a gene list ranked by a metric that captures the magnitude of differential expression. Here we use the DESeq2 log2 fold change (WBH vs. Sham), sorted in decreasing order so that the most upregulated genes appear at the top of the list.

```{r ranked-gene-list}
gsea_ranked_log2fc <- res_df %>%
  dplyr::filter(!is.na(log2FoldChange)) %>%
  dplyr::select(gene_id, log2FoldChange) %>%
  dplyr::arrange(desc(log2FoldChange))

head(gsea_ranked_log2fc)
```

## Define gene sets for GSEA from GO IDs

Gene sets are assembled by querying `org.Hs.eg.db` for Biological Process GO terms whose descriptions match four keywords of interest: *heat/chaperone/protein folding/proteostasis*, *interferon*, *serotonin*, and *dopamine*. Only genes present in the ranked list are retained, giving four gene sets of sizes ranging from 46 (serotonin) to 374 (interferon signaling).

```{r define-gene-sets}
# Create named ranked vector for GSEA (gene -> statistic)
ranked_stats        <- gsea_ranked_log2fc$log2FoldChange
names(ranked_stats) <- gsea_ranked_log2fc$gene_id
ranked_stats        <- sort(ranked_stats, decreasing = TRUE)

# Map gene symbols to GO terms (Biological Process only)
go_map <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys    = keys(org.Hs.eg.db, keytype = "SYMBOL"),
  columns = c("SYMBOL", "GO", "ONTOLOGY"),
  keytype = "SYMBOL"
) %>%
  filter(!is.na(GO), ONTOLOGY == "BP")

# Retrieve GO term names (Biological Process ontology)
go_terms <- AnnotationDbi::select(
  GO.db,
  keys    = keys(GO.db),
  columns = c("TERM", "ONTOLOGY"),
  keytype = "GOID"
) %>%
  dplyr::filter(ONTOLOGY == "BP")

# Create mapping: GO ID -> gene symbols
go2gene <- split(go_map$SYMBOL, go_map$GO)

# Identify relevant GO terms by keyword matching
heat_go_ids <- go_terms %>%
  filter(str_detect(
    TERM,
    regex("heat|chaperone|protein folding|protein refolding|proteostasis",
          ignore_case = TRUE)
  )) %>%
  pull(GOID)

interferon_go_ids <- go_terms %>%
  filter(str_detect(TERM, regex("interferon", ignore_case = TRUE))) %>%
  pull(GOID)

serotonin_go_ids <- go_terms %>%
  filter(str_detect(TERM, regex("serotonin", ignore_case = TRUE))) %>%
  pull(GOID)

dopamine_go_ids <- go_terms %>%
  filter(str_detect(TERM, regex("dopamine", ignore_case = TRUE))) %>%
  pull(GOID)

# Build gene sets from selected GO terms
gene_sets <- list(
  heat_shock_proteostasis = unique(unlist(go2gene[heat_go_ids])),
  interferon_signaling    = unique(unlist(go2gene[interferon_go_ids])),
  serotonin               = unique(unlist(go2gene[serotonin_go_ids])),
  dopamine                = unique(unlist(go2gene[dopamine_go_ids]))
)

# Keep only genes present in the ranked list
gene_sets <- lapply(gene_sets, function(gs) {
  intersect(unique(gs), names(ranked_stats))
})

# Summarize gene set sizes after filtering
gene_set_summary <- tibble(
  gene_set               = names(gene_sets),
  n_genes_in_ranked_list = sapply(gene_sets, length)
)
gene_set_summary
```

## Define Enrichment Score function

The enrichment score (ES) is computed using the weighted Kolmogorov–Smirnov-like statistic from the original Subramanian et al. (2005) paper. At each position in the ranked gene list, the running sum increases by the gene's absolute fold-change (weighted hit) if the gene belongs to the set, and decreases by an equal penalty otherwise. The ES is the maximum absolute deviation of this running sum from zero.

```{r es-function}
gsea_es <- function(ranked_stats, gene_set, p = 1) {
  genes <- names(ranked_stats)
  N     <- length(genes)
  hits  <- genes %in% gene_set
  Nh    <- sum(hits)

  if (Nh == 0) return(NA)

  # Compute weights
  weights <- abs(ranked_stats)^p

  # Normalization constants
  NR <- sum(weights[hits])

  # Initialize cumulative vectors
  Phit  <- numeric(N)
  Pmiss <- numeric(N)

  # Compute cumulative sums explicitly
  hit_cum  <- 0
  miss_cum <- 0

  for (i in seq_len(N)) {
    if (hits[i]) {
      hit_cum  <- hit_cum  + weights[i] / NR
    } else {
      miss_cum <- miss_cum + 1 / (N - Nh)
    }
    Phit[i]  <- hit_cum
    Pmiss[i] <- miss_cum
  }

  running_score <- Phit - Pmiss

  # Enrichment score = max absolute deviation
  ES <- running_score[which.max(abs(running_score))]

  return(list(
    ES            = ES,
    running_score = running_score,
    Phit          = Phit,
    Pmiss         = Pmiss
  ))
}
```

## Permutation Test for Normalized Enrichment Score (NES) and P-value

Statistical significance is assessed via phenotype permutation. Rather than refitting the full DESeq2 model 1,000 times (which would be prohibitively slow), we permute sample labels and recompute a fast ranking statistic: the difference in mean log-CPM between the permuted WBH and Sham groups. The observed ES for each gene set is then normalized by the mean of the same-sign null ES values to produce the NES, and an empirical p-value is derived from the fraction of permutations with a null |ES| greater than or equal to the observed |ES|. Multiple testing correction uses both Benjamini–Hochberg (BH) and Storey's q-value.

```{r permutation-test}
# phenotype permutation using logCPM mean differences
logcpm <- edgeR::cpm(dge, log = TRUE, prior.count = 1)

rank_from_labels <- function(logcpm, group) {
  group <- factor(group, levels = c("Sham", "WBH"))
  stat  <- rowMeans(logcpm[, group == "WBH",  drop = FALSE]) -
           rowMeans(logcpm[, group == "Sham", drop = FALSE])
  sort(stat, decreasing = TRUE)
}

compute_NES <- function(observed_ES, null_dist) {
  if (is.na(observed_ES)) return(NA_real_)
  if (observed_ES >= 0) {
    pos_null <- null_dist[null_dist >= 0]
    if (length(pos_null) == 0 || mean(pos_null) == 0) return(NA_real_)
    observed_ES / mean(pos_null)
  } else {
    neg_null <- null_dist[null_dist < 0]
    if (length(neg_null) == 0 || mean(abs(neg_null)) == 0) return(NA_real_)
    observed_ES / mean(abs(neg_null))
  }
}

gsea_pheno_perm <- function(dge, gene_sets, observed_ranked_stats, nperm = 1000) {
  logcpm <- edgeR::cpm(dge, log = TRUE, prior.count = 1)

  observed_ES <- sapply(gene_sets, function(gs) {
    gsea_es(observed_ranked_stats, gs)$ES
  })

  null_mat <- matrix(
    NA_real_,
    nrow     = nperm,
    ncol     = length(gene_sets),
    dimnames = list(NULL, names(gene_sets))
  )

  for (b in seq_len(nperm)) {
    perm_group         <- sample(dge$samples$group)
    perm_ranked_stats  <- rank_from_labels(logcpm, perm_group)
    null_mat[b, ]      <- sapply(gene_sets, function(gs) {
      gsea_es(perm_ranked_stats, gs)$ES
    })
  }

  pvalues <- sapply(seq_along(gene_sets), function(j) {
    mean(abs(null_mat[, j]) >= abs(observed_ES[j]), na.rm = TRUE)
  })

  NES <- sapply(seq_along(gene_sets), function(j) {
    compute_NES(
      observed_ES = observed_ES[j],
      null_dist   = null_mat[, j]
    )
  })

  gsea_df <- tibble(
    gene_set = names(gene_sets),
    ES       = observed_ES,
    NES      = NES,
    pvalue   = pvalues,
    padj_BH  = p.adjust(pvalues, method = "BH")
  ) %>%
    arrange(pvalue)

  lambda    <- 0.5
  pi0_storey <- min(1, mean(gsea_df$pvalue > lambda, na.rm = TRUE) / (1 - lambda))

  gsea_df <- gsea_df %>%
    mutate(
      pi0_storey    = pi0_storey,
      qvalue_storey = pi0_storey * pvalue
    )

  return(gsea_df)
}

set.seed(114)
gsea_df <- gsea_pheno_perm(
  dge                  = dge,
  gene_sets            = gene_sets,
  observed_ranked_stats = ranked_stats,
  nperm                = 1000
)
```

## GSEA Results

None of the four gene sets reach statistical significance after multiple-testing correction (all BH-adjusted p-values >= 0.844). The heat shock/proteostasis set shows the most positive NES (1.103), consistent with the expected cellular response to elevated temperature, while the dopamine, serotonin, and interferon sets all have negative NES values, suggesting slight depletion relative to chance.


```{r gsea-results}
gene_set_labels <- c(
  heat_shock_proteostasis = "Heat Shock Proteasis",
  dopamine                = "Dopamine",
  serotonin               = "Serotonin",
  interferon_signaling    = "Interferon Signaling"
)

gsea_df <- gsea_df %>%
  mutate(
    gene_set = recode(gene_set, !!!gene_set_labels)
  )

gsea_table <- gsea_df |>
  gt() |>
  tab_header(
    title    = "Gene Set Enrichment Analysis Results",
    subtitle = "Summary of enrichment scores and significance metrics"
  ) |>
  cols_label(
    gene_set      = "Gene Set",
    ES            = "ES",
    NES           = "NES",
    pvalue        = "P-value",
    padj_BH       = "Adj. P (BH)",
    pi0_storey    = "Pi0 (Storey)",
    qvalue_storey = "Q-value"
  ) |>
  fmt_number(columns = c(ES, NES), decimals = 3) |>
  fmt_number(columns = c(pvalue, padj_BH, qvalue_storey), decimals = 3) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) |>
  cols_align(align = "center", columns = -gene_set) |>
  tab_options(
    table.font.size     = 10,
    table.width         = pct(100),
    data_row.padding    = px(3),
    column_labels.padding = px(3)
  )

gsea_table
```

## Plotting Enrichment Scores

The bar chart below displays the absolute NES for each gene set, colored by its BH-adjusted p-value. Redder bars indicate lower (more significant) adjusted p-values, while bluer bars indicate higher (less significant) values. 

```{r bar-plot, fig.height=4}
ggplot(gsea_df, aes(x = gene_set, y = abs(NES), fill = padj_BH)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(
    low  = "red",
    high = "blue",
    name = "BH adjusted\np-value"
  ) +
  labs(
    title = "GSEA Results by Gene Set",
    x     = "Gene set",
    y     = "Normalized Enrichment Score"
  ) +
  theme_minimal()
```

The running enrichment score plots below show, for each gene set, how the cumulative sum evolves as we walk down the ranked gene list from most upregulated (left) to most downregulated (right). Tick marks on the x-axis indicate where gene-set members appear in the list; the dotted vertical line marks the position of the maximum absolute deviation (the reported ES).

```{r es-plot-function}
plot_gsea_es <- function(ranked_stats, gene_set, gene_set_name) {
  es <- gsea_es(ranked_stats, gene_set)

  plot(
    es$running_score,
    type = "l",
    lwd  = 2,
    col  = "green",
    xlab = "Gene Rank in Ordered List",
    ylab = "Running Enrichment Score (ES)",
    main = paste("GSEA Enrichment Score:", gene_set_name)
  )
  abline(h = 0, lty = 2)

  hit_positions <- which(names(ranked_stats) %in% gene_set)
  rug(hit_positions)

  abline(
    v   = which.max(abs(es$running_score)),
    lty = 3
  )
}
```

### Heat Shock / Proteostasis

The heat shock gene set (293 genes) shows a broad positive enrichment: the running score climbs steadily through the upper half of the ranked list before falling back, with its peak (ES = 0.215) occurring around rank 8,000. This pattern is consistent with mild but diffuse upregulation of chaperones and proteostasis genes following WBH treatment.

```{r es-heat, fig.height=4.5}
plot_gsea_es(
  ranked_stats  = ranked_stats,
  gene_set      = gene_sets[["heat_shock_proteostasis"]],
  gene_set_name = gene_set_labels[["heat_shock_proteostasis"]]
)
```

### Interferon Signaling

The interferon signaling set (374 genes) produces a small positive peak early in the list before a sustained decline, yielding a negative final ES (-0.153). This suggests that interferon pathway genes are not preferentially upregulated by WBH.

```{r es-interferon, fig.height=4.5}
plot_gsea_es(
  ranked_stats  = ranked_stats,
  gene_set      = gene_sets[["interferon_signaling"]],
  gene_set_name = gene_set_labels[["interferon_signaling"]]
)
```

### Serotonin

The serotonin gene set (46 genes) shows a very early, brief positive peak followed by a nearly linear decline through the remainder of the list, resulting in an ES of -0.463. The plot reflects the small set size and the presence serotonin genes across the lower half of the ranking.

```{r es-serotonin, fig.height=4.5}
plot_gsea_es(
  ranked_stats  = ranked_stats,
  gene_set      = gene_sets[["serotonin"]],
  gene_set_name = gene_set_labels[["serotonin"]]
)
```

### Dopamine

Dopamine pathway genes (131 genes) show a pattern similar to serotonin: a small positive early peak followed by a long linear decline, giving the largest negative ES of the four sets (-0.541). Taken together with the serotonin result, this suggests that neurotransmitter signaling genes tend to be downregulated, or at least not upregulated under WBH in this dataset.

```{r es-dopamine, fig.height=4.5}
plot_gsea_es(
  ranked_stats  = ranked_stats,
  gene_set      = gene_sets[["dopamine"]],
  gene_set_name = gene_set_labels[["dopamine"]]
)
```

## Session info

```{r session-info}
sessionInfo()
```
